\bf Tearing Mode Dynamics in Tokamak Plasmas

### 2021-11-01 Documentation/TearingModeTheory.tex
Added to tearing writeup

\section{Fluid Theory}
\subsection{Introduction}
The aim of this section is to introduce the fundamental plasma fluid theory that underpins the analysis of tearing mode dynamics in tokamak
plasmas.

@@

\subsection{Braginskii Equations}\label{sbrag}
The electron and ion fluid equations in a collisional plasma
take the form:
\begin{align}\label{e3.79a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e \,n\,\frac{d {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F},\label{e3.79b}\\[0.5ex]
\frac{3}{2}\frac{d p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_e &= w_e,\label{e3.79c}
\end{align}
and
\begin{align}\label{e3.80a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i \,n\,\frac{d {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F},\label{e3.80b}\\[0.5ex]
\frac{3}{2}\frac{d p_i}{dt} + \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i &= w_i,\label{e3.80c}
\end{align}
respectively. 
Equations (\ref{e3.79a})--(\ref{e3.80c})  are called the {\em Braginskii equations}, because they were first obtained
in a celebrated article by S.I.~Braginskii (Braginskii 1965). Here, $n$ is the electron number density, ${\bf V}_s$ is the species-$s$ flow velocity, $p_s=n\,T_s$ is the 
species-$s$ scalar
pressure, $\bpi_s$ is the species-$s$ viscosity tensor, ${\bf F}$ is the collisional friction force density, ${\bf q}_s$ is the species-$s$ heat flux density, $w_s$ is the species-$s$ collisional heating rate density, and ${\bf E}$ is the ambient electric field strength. Moreover, $d/dt\equiv \partial/\partial t + {\bf V}_e\cdot\nabla$ in the electron fluid equations, (\ref{e3.79a})--(\ref{e3.79c}), whereas  $d/dt\equiv \partial/\partial t + {\bf V}_i\cdot\nabla$ in the ion fluid equations,  (\ref{e3.80a})--(\ref{e3.80c}).

@@

 It is convenient to split the friction force
${\bf F}$ into a component ${\bf F}_U$ corresponding to resistivity, and a
component ${\bf F}_T$ corresponding to the thermal force. Thus,
\begin{equation}
{\bf F} = {\bf F}_U+{\bf F}_T,
\end{equation}
where
\begin{align}
{\bf F}_U&= n\,e\left(\frac{{\bf j}_\parallel}{\sigma_\parallel}
+\frac{{\bf j}_\perp}{\sigma_\perp}\right), \\[0.5ex]
{\bf F}_T &= - 0.71\,n\,\nabla_\parallel T_e
-\frac{3\,n}{2\,{\mit\Omega}_e\,\tau_e}\,{\bf b}\times\nabla_\perp T_e.
\end{align}
Likewise, the electron collisional energy gain term $w_e$ is split
into a component $-w_i$ corresponding to the energy lost to the ions (in the
ion rest frame), a component $w_U$ corresponding to work done by the friction
force ${\bf F}_U$, and a component $w_T$ corresponding to work done by the
thermal force ${\bf F}_T$. Thus,
\begin{equation}
w_e = -w_i + w_U + w_T,
\end{equation}
where
\begin{align}
w_U &= \frac{{\bf j}\cdot{\bf F}_U}{n\,e},\\[0.5ex]
w_T&= \frac{{\bf j}\cdot{\bf F}_T}{n\,e}.
\end{align}
Finally, it is helpful to split the electron heat flux density ${\bf q}_e$ into
a diffusive component ${\bf q}_{Te}$ and a convective component ${\bf q}_{Ue}$. 
Thus,
\begin{equation}
{\bf q}_e = {\bf q}_{Te} + {\bf q}_{Ue},
\end{equation}
where
\begin{align}
{\bf q}_{Te} &=-\kappa_\parallel^e\,\nabla_\parallel T_e -\kappa_\perp^e\,
\nabla_\perp T_e
-\kappa_\times^e\,{\bf b}\times\nabla_\perp T_e,\\[0.5ex]
{\bf q}_{Ue}&=  0.71\,\frac{T_e}{e}\,{\bf j}_\parallel-
\frac{3\,T_e}{2\,{\mit\Omega}_e\,\tau_e\,e}\,{\bf b}\times{\bf j}_\perp.
\end{align}

Let us, first of all, consider the electron fluid equations, which can
be written:
\begin{align}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e\, n\,\frac{d {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F}_U +{\bf F}_T,\\[0.5ex]
\frac{3}{2}\frac{d p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_{Te} 
+ \nabla\cdot{\bf q}_{Ue} &= -w_i+w_U + w_T.
\end{align}
Let $\bar{n}$, $\bar{v}_e$, $\bar{\tau}_e$, $\bar{l}_e=\bar{v}_e\,\bar{\tau}_e$, $\bar{B}$,
and $\bar{\rho}_e =\bar{ v}_e/(e\bar{B}/m_e)$,  be typical values
of the particle density, the electron thermal velocity, the electron collision time, the electron
mean-free-path between collisions,  the magnetic field-strength, and the
electron gyroradius,  respectively. 
Suppose that the typical spatial variation lengthscale of fluid variables is $L$. Let
 \begin{align}
\delta_e &=\frac{\bar{\rho}_e}{L},\\[0.5ex]
\zeta_e &= \frac{\bar{\rho}_e}{\bar{l}_e},\\[0.5ex]
\mu &= \sqrt{\frac{m_e}{m_i}}.
\end{align}
All three of these parameters are assumed to be  small
compared to unity. Finally, the typical electron flow velocity is assumed to be of order $\delta_e\,\bar{v}_e$. This corresponds
to the so-called {\em drift ordering}\/ in which the flow velocity is comparable to the curvature and grad-B particle
drift velocities (Fitzpatrick 2015). The drift ordering is appropriate to tearing modes in tokamak plasmas, which are comparatively
slowly growing instabilities.

@@

The normalized electron fluid equations take the form:
\begin{align}
\frac{d\hat{n}}{d\hat{t}} + \hat{n}\,\widehat{\nabla}\cdot\widehat{\bf V}_e &=0,\label{e3.113a}\\[0.5ex]
\hat{n}\,\frac{d\widehat{\bf V}_e}{d\hat{t}} +
\delta_e^{-2}\,\widehat{\nabla}\hat{p}_e +\zeta_e^{-1}\,\widehat{\nabla}
\cdot\widehat{\bpi}_e  
+ \delta_e^{-2}\,\hat{n}\,(\widehat{\bf E} + \widehat{\bf V}_e
\times\widehat{\bf B}) &= \delta_e^{-2}\,\zeta_e\,\widehat{\bf F}_U +\delta_e^{-2}\,\widehat{\bf F}_T, \\[0.5ex]
\frac{3}{2}\frac{d\hat{p}_e}{d\hat{t}} + \frac{5}{2}\,
\hat{p}_e\,\widehat{\nabla}\cdot\widehat{\bf V}_e + \delta_e^{2}\,
\zeta_e^{-1}\,\widehat{\bpi}_e : \widehat{\nabla}\cdot\widehat{\bf V}_e&\nonumber\label{e3.113c}\\[0.5ex] 
+\zeta_e^{-1} \,\widehat{\nabla}\cdot\widehat{\bf q}_{Te}
 +\widehat{\nabla}\cdot\widehat{\bf q}_{Ue} 
&= -\delta_e^{-2}\,\zeta_e\,\mu^2\,\widehat{w}_i + \zeta_e\,
\widehat{w}_U +\widehat{w}_T.
\end{align}
The only large or small (compared to unity) quantities in these equations are the
parameters $\delta_e$, $\zeta_e$, and $\mu$. 
Here, $d/d\hat{t}\equiv\partial /\partial\hat{t} +
 \widehat{\bf V}_e\cdot\widehat{\nabla}$. It is assumed that $T_e\sim T_i$.

Let us now consider the ion fluid equations, which can be written:
\begin{align}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i\, n\,\frac{d {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F}_U -{\bf F}_T,\\[0.5ex]
\frac{3}{2}\frac{d p_i}{dt} + \frac{5}{2}\,p_i\,\nabla \cdot {\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i & =w_i.
\end{align}
It is convenient to adopt a normalization scheme for the ion equations
which is similar to, but independent of, that employed to normalize the
electron equations. Let  $\bar{n}$, $\bar{v}_i$, $\bar{l}_i$, $\bar{B}$,
and $\bar{\rho}_i =\bar{ v}_i/(e\bar{B}/m_i)$,  be typical values
of the particle density, the ion  thermal velocity, the ion
mean-free-path between collisions,  the magnetic field-strength, and the
ion gyroradius,  respectively. 
Suppose that 
the typical spatial variation lengthscale of fluid quantities is $L$. Let
 \begin{align}
\delta_i &= \frac{\bar{\rho}_i}{L},\\[0.5ex]
\zeta_i &= \frac{\bar{\rho}_i}{\bar{l}_i},\\[0.5ex]
\mu &= \sqrt{\frac{m_e}{m_i}}.
\end{align}
All three of these parameters are assumed to be  small 
compared to unity.  As before, we adopt the drift ordering in which the typical ion flow velocity is assumed to be of order $\delta_i\,\bar{v}_i$.

@@

The normalized ion fluid equations take the form:
\begin{align}\label{e3.116a}
\frac{d\hat{n}}{d\hat{t}} + \hat{n}\,\widehat{\nabla}\cdot\widehat{\bf V}_i &=0,\\[0.5ex]
\hat{n}\,\frac{d\widehat{\bf V}_i}{d\hat{t}} +
\delta_i^{-2}\,\widehat{\nabla}\hat{p}_i+\zeta_i^{-1}\,\widehat{\nabla}
\cdot\widehat{\bpi}_i  
-\delta_i^{-2}\,\hat{n}\,(\widehat{\bf E} + \widehat{\bf V}_i
\times\widehat{\bf B}) &=-\delta_i^{-2}\,\zeta_i\,\mu\,\widehat{\bf F}_U
 - \delta_i^{-2}\,\widehat{\bf F}_T, \\[0.5ex]
\frac{3}{2}\frac{d\hat{p}_i}{d\hat{t}} + \frac{5}{2}\,
\hat{p}_i\,\widehat{\nabla}\cdot\widehat{\bf V}_i + \delta_i^{2}\,
\zeta_i^{-1}\,\widehat{\bpi}_i : \widehat{\nabla}\cdot\widehat{\bf V}_i
+\zeta_i^{-1} \,\widehat{\nabla}\cdot\widehat{\bf q}_{i}
&= \delta_i^{-2}\,\zeta_i\,\mu\,\widehat{w}_i. \label{e3.116c}
\end{align}
The only large or small (compared to unity) quantities in these equations are the
parameters $\delta_i$, $\zeta_i$, and $\mu$. 
Here, $d/d\hat{t}\equiv\partial /\partial\hat{t} +
 \widehat{\bf V}_i\cdot\widehat{\nabla}$.

@@

\subsection{Electromagnetic Torques}\label{storque}
The flux-surface integrated poloidal and toroidal electromagnetic torque densities acting on the plasma
can be written
\begin{align}\label{e152}
T_\theta(r) &= \left\langle r\,{\bf j}\times {\bf B}\cdot {\bf e}_\theta\right\rangle\\[0.5ex]
T_\varphi(r) &= \left\langle R_0\,{\bf j}\times {\bf B}\cdot {\bf e}_z\right\rangle,
\end{align}
respectively, 
where
\begin{equation}
\left\langle\cdots\right\rangle \equiv \oint\oint r\,R_0\,(\cdots)\,d\theta\,d\varphi
\end{equation}
is a flux-surface integration  operator. 
 However, according to Equation~(\ref{e80}),
\begin{equation}\label{e150}
{\bf j}\times {\bf B} \simeq \nabla p.
\end{equation}
Given that the scalar pressure is a single-valued function of $\theta$ and $\varphi$, it immediately follows
that $T_\theta=T_\varphi = 0$ throughout the plasma (Fitzpatrick 1993). The only exception to this rule occurs in the immediate vicinity of the
resonant surface, where Equation~(\ref{e150}) breaks down. It follows that we can write
\begin{align}\label{e156a}
T_\theta(r) &= T_{\theta\,s}\,\delta(r-r_s),\\[0.5ex]
T_\varphi(r) &= T_{\varphi\,s}\,\delta(r-r_s),
\end{align}
where
\begin{align}\label{e153}
T_{\theta\,s} &=  \frac{1}{4}\int_{r_{s-}}^{r_{s+}}\oint\oint R_0\,r^{\,2}\,(\delta j_z\,\delta B_r^{\,\ast} + \delta j_z^{\,\ast}\,\delta B_r)\,dr\,d\theta\,d\varphi,\\[0.5ex]
T_{\varphi\,s} &=  -\frac{1}{4}\int_{r_{s-}}^{r_{s+}}\oint\oint R_0^{\,2}\,r\,(\delta j_\theta\,\delta B_r^{\,\ast} + \delta j_\theta^{\,\ast}\,\delta B_r)\,dr\,d\theta\,d\varphi
\end{align}
are the net poloidal and toroidal torques, respectively, acting at the resonant surface. 
Note that the zeroth-order (in perturbed quantities) torques are zero because $B_r=0$. Furthermore, the linear (in perturbed quantities) torques 
average to zero over the flux-surface. Hence, the largest non-zero torques  are quadratic in perturbed quantities. 
According to Equation~(\ref{e82}), 
\begin{equation}
\nabla \cdot \delta {\bf j} = {\rm i}\left(\frac{m}{r}\,\delta j_\theta - \frac{n}{R_0}\,\delta j_z\right) = 0,
\end{equation}
where $\delta{\bf j}$ is the density of the current sheet flowing at the resonant surface, and we have made use
of the fact that $\delta j_r\simeq 0$. 
 It follows from the previous three equations that
\begin{equation}\label{e156}
T_{\varphi\,s} = - \frac{n}{m}\,T_{\theta\,s}.
\end{equation}
Finally, Equations~(\ref{e93}), (\ref{e94}), (\ref{e153}), and (\ref{e156}) imply that
\begin{align}\label{e157}
T_{\theta\,s} &= -\frac{2\pi^2\,R_0\,m}{\mu_0}\,{\rm Im}({\mit\Delta\Psi}_s\,{\mit\Psi}_s^{\,\ast}),\\[0.5ex]
T_{\varphi\,s} &= \frac{2\pi^2\,R_0\,n}{\mu_0}\,{\rm Im}({\mit\Delta\Psi}_s\,{\mit\Psi}_s^{\,\ast})\label{e158}
\end{align}
(Fitzpatrick 1993).

@@

Equations~(\ref{e182}) and (\ref{e185}) are subject to the boundary conditions
\begin{align}\label{e186}
\frac{\partial{\mit\Delta\Omega}_\theta(0,t)}{\partial r} &= \frac{\partial{\mit\Delta\Omega}_\varphi(0,t)}{\partial r} =0,\\[0.5ex]
{\mit\Delta\Omega}_\theta(a,t) &= {\mit\Delta\Omega}_\varphi(a,t)=0.\label{e187}
\end{align}
The boundary conditions (\ref{e186}) merely ensure that the ion angular velocities remain finite at the magnetic axis. 
On the other  hand, the boundary conditions (\ref{e187}) are a consequence of the action of charge exchange with neutrals emitted isotropically from the wall  in the edge regions of the plasma
(Brau, et al., 1983; Fitzpatrick 1993; Monier-Garbet, et al., 1997). Charge exchange with neutrals gives rise to 
dominant damping torques acting at the edge of the plasma that relaxes the edge ion angular velocities  toward  particular values. Moreover, the
electromagnetic torques that develop at the resonant surface are not large enough, compared with the charge-exchange torques,
to modify the edge ion angular velocities (Fitzpatrick 1993).

@@

Equations~(\ref{e220})---(\ref{e223}) are known collectively as the {\em drift-MHD equations}, and contain both shear-Alfv\'{e}n
and compressible-Alfv\'{e}n wave dynamics. The system of equations is completed by Maxwell's equations:
\begin{align}
\nabla\cdot{\bf B} &= 0,\\[0.5ex]
\nabla\times {\bf E} &= - \frac{\partial {\bf B}}{\partial t},\\[0.5ex]
\mu_0\,{\bf j} &= \nabla \times {\bf B}.\label{e226}
\end{align}

>>>>

\section{Introduction}

\section{Plasma Fluid Theory}
\subsection{Introduction}
The aim of this section is to introduce the fundamental plasma fluid theory that underpins the analysis of tearing mode dynamics in tokamak
plasmas.

@@

\subsection{Braginskii Equations}\label{sbrag}
The electron and ion fluid equations in a collisional plasma
take the form:
\begin{align}\label{e3.79a}
\frac{d_en}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e \,n\,\frac{d_e {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F},\label{e3.79b}\\[0.5ex]
\frac{3}{2}\frac{d_e p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_e &= w_e,\label{e3.79c}
\end{align}
and
\begin{align}\label{e3.80a}
\frac{d_i n}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i \,n\,\frac{d_i {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F},\label{e3.80b}\\[0.5ex]
\frac{3}{2}\frac{d_i p_i}{dt} + \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i &= w_i,\label{e3.80c}
\end{align}
respectively. 
Equations (\ref{e3.79a})--(\ref{e3.80c})  are called the {\em Braginskii equations}, because they were first obtained
in a celebrated article by S.I.~Braginskii (Braginskii 1965). Here, $n$ is the electron number density, ${\bf V}_s$ is the species-$s$ flow velocity, $p_s=n\,T_s$ is the 
species-$s$ scalar
pressure, $\bpi_s$ is the species-$s$ viscosity tensor, ${\bf F}$ is the collisional friction force density, ${\bf q}_s$ is the species-$s$ heat flux density, $w_s$ is the species-$s$ collisional heating rate density, and ${\bf E}$ is the ambient electric field-strength. Moreover, $d_e/dt\equiv \partial/\partial t + {\bf V}_e\cdot\nabla$ in the electron fluid equations, (\ref{e3.79a})--(\ref{e3.79c}), whereas  $d_i/dt\equiv \partial/\partial t + {\bf V}_i\cdot\nabla$ in the ion fluid equations,  (\ref{e3.80a})--(\ref{e3.80c}).

@@

 It is convenient to split the friction force density
${\bf F}$ into a component ${\bf F}_U$ corresponding to resistivity, and a
component ${\bf F}_T$ corresponding to the so-called {\em thermal force}. Thus,
\begin{equation}
{\bf F} = {\bf F}_U+{\bf F}_T,
\end{equation}
where
\begin{align}
{\bf F}_U&= n\,e\left(\frac{{\bf j}_\parallel}{\sigma_\parallel}
+\frac{{\bf j}_\perp}{\sigma_\perp}\right), \\[0.5ex]
{\bf F}_T &= - 0.71\,n\,\nabla_\parallel T_e
-\frac{3\,n}{2\,{\mit\Omega}_e\,\tau_e}\,{\bf b}\times\nabla_\perp T_e.
\end{align}
Likewise, the electron collisional energy gain term $w_e$ is split
into a component $-w_i$ corresponding to the energy lost to the ions (in the
ion rest frame), a component $w_U$ corresponding to work done by the friction
force ${\bf F}_U$, and a component $w_T$ corresponding to work done by the
thermal force ${\bf F}_T$. Thus,
\begin{equation}
w_e = -w_i + w_U + w_T,
\end{equation}
where
\begin{align}
w_U &= \frac{{\bf j}\cdot{\bf F}_U}{n\,e},\\[0.5ex]
w_T&= \frac{{\bf j}\cdot{\bf F}_T}{n\,e}.
\end{align}
Finally, it is helpful to split the electron heat flux density ${\bf q}_e$ into
a diffusive component ${\bf q}_{Te}$ and a convective component ${\bf q}_{Ue}$. 
Thus,
\begin{equation}
{\bf q}_e = {\bf q}_{Te} + {\bf q}_{Ue},
\end{equation}
where
\begin{align}
{\bf q}_{Te} &=-\kappa_\parallel^e\,\nabla_\parallel T_e -\kappa_\perp^e\,
\nabla_\perp T_e
-\kappa_\times^e\,{\bf b}\times\nabla_\perp T_e,\\[0.5ex]
{\bf q}_{Ue}&=  0.71\,\frac{T_e}{e}\,{\bf j}_\parallel-
\frac{3\,T_e}{2\,{\mit\Omega}_e\,\tau_e\,e}\,{\bf b}\times{\bf j}_\perp.
\end{align}

Let us, first of all, consider the electron fluid equations, which can
be written:
\begin{align}
\frac{d_e n}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e\, n\,\frac{d_e {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F}_U +{\bf F}_T,\\[0.5ex]
\frac{3}{2}\frac{d_e p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_{Te} 
+ \nabla\cdot{\bf q}_{Ue} &= -w_i+w_U + w_T.
\end{align}
Let $\bar{n}$, $\bar{v}_e$, $\bar{\tau}_e$, $\bar{l}_e=\bar{v}_e\,\bar{\tau}_e$, $\bar{B}$,
and $\bar{\rho}_e =\bar{ v}_e/(e\bar{B}/m_e)$,  be typical values
of the particle density, the electron thermal velocity, the electron collision time, the electron
mean-free-path between collisions,  the magnetic field-strength, and the
electron gyroradius,  respectively. 
Suppose that the typical spatial variation lengthscale of fluid variables is $L$. Let
 \begin{align}
\delta_e &=\frac{\bar{\rho}_e}{L},\\[0.5ex]
\zeta_e &= \frac{\bar{\rho}_e}{\bar{l}_e},\\[0.5ex]
\mu &= \sqrt{\frac{m_e}{m_i}}.
\end{align}
All three of these parameters are assumed to be  small
compared to unity. Finally, the typical electron flow velocity is assumed to be of order $\delta_e\,\bar{v}_e$. This corresponds
to the so-called {\em drift ordering}\/ in which the flow velocity is comparable to the curvature and grad-B particle
drift velocities (Fitzpatrick 2015). The drift ordering is appropriate to tearing modes in tokamak plasmas, which are comparatively
slowly growing instabilities.

@@

The normalized electron fluid equations take the form:
\begin{align}
\frac{d_e \hat{n}}{d\hat{t}} + \hat{n}\,\widehat{\nabla}\cdot\widehat{\bf V}_e &=0,\label{e3.113a}\\[0.5ex]
\hat{n}\,\frac{d_e \widehat{\bf V}_e}{d\hat{t}} +
\delta_e^{-2}\,\widehat{\nabla}\hat{p}_e +\zeta_e^{-1}\,\widehat{\nabla}
\cdot\widehat{\bpi}_e  
+ \delta_e^{-2}\,\hat{n}\,(\widehat{\bf E} + \widehat{\bf V}_e
\times\widehat{\bf B}) &= \delta_e^{-2}\,\zeta_e\,\widehat{\bf F}_U +\delta_e^{-2}\,\widehat{\bf F}_T, \\[0.5ex]
\frac{3}{2}\frac{d_e \hat{p}_e}{d\hat{t}} + \frac{5}{2}\,
\hat{p}_e\,\widehat{\nabla}\cdot\widehat{\bf V}_e + \delta_e^{2}\,
\zeta_e^{-1}\,\widehat{\bpi}_e : \widehat{\nabla}\cdot\widehat{\bf V}_e&\nonumber\label{e3.113c}\\[0.5ex] 
+\zeta_e^{-1} \,\widehat{\nabla}\cdot\widehat{\bf q}_{Te}
 +\widehat{\nabla}\cdot\widehat{\bf q}_{Ue} 
&= -\delta_e^{-2}\,\zeta_e\,\mu^2\,\widehat{w}_i + \zeta_e\,
\widehat{w}_U +\widehat{w}_T.
\end{align}
The only large or small (compared to unity) quantities in these equations are the
parameters $\delta_e$, $\zeta_e$, and $\mu$. 
Here, $d_e/d\hat{t}\equiv\partial /\partial\hat{t} +
 \widehat{\bf V}_e\cdot\widehat{\nabla}$. It is assumed that $T_e\sim T_i$.

Let us now consider the ion fluid equations, which can be written:
\begin{align}
\frac{d_i n}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i\, n\,\frac{d_i {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F}_U -{\bf F}_T,\\[0.5ex]
\frac{3}{2}\frac{d_i p_i}{dt} + \frac{5}{2}\,p_i\,\nabla \cdot {\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i & =w_i.
\end{align}
It is convenient to adopt a normalization scheme for the ion equations
which is similar to, but independent of, that employed to normalize the
electron equations. Let  $\bar{n}$, $\bar{v}_i$, $\bar{l}_i$, $\bar{B}$,
and $\bar{\rho}_i =\bar{ v}_i/(e\bar{B}/m_i)$,  be typical values
of the particle density, the ion  thermal velocity, the ion
mean-free-path between collisions,  the magnetic field-strength, and the
ion gyroradius,  respectively. 
Suppose that 
the typical spatial variation lengthscale of fluid quantities is $L$. Let
 \begin{align}
\delta_i &= \frac{\bar{\rho}_i}{L},\\[0.5ex]
\zeta_i &= \frac{\bar{\rho}_i}{\bar{l}_i},\\[0.5ex]
\mu &= \sqrt{\frac{m_e}{m_i}}.
\end{align}
All three of these parameters are assumed to be  small 
compared to unity.  As before, we adopt the drift ordering in which the typical ion flow velocity is assumed to be of order $\delta_i\,\bar{v}_i$.

@@

The normalized ion fluid equations take the form:
\begin{align}\label{e3.116a}
\frac{d_i \hat{n}}{d\hat{t}} + \hat{n}\,\widehat{\nabla}\cdot\widehat{\bf V}_i &=0,\\[0.5ex]
\hat{n}\,\frac{d_i \widehat{\bf V}_i}{d\hat{t}} +
\delta_i^{-2}\,\widehat{\nabla}\hat{p}_i+\zeta_i^{-1}\,\widehat{\nabla}
\cdot\widehat{\bpi}_i  
-\delta_i^{-2}\,\hat{n}\,(\widehat{\bf E} + \widehat{\bf V}_i
\times\widehat{\bf B}) &=-\delta_i^{-2}\,\zeta_i\,\mu\,\widehat{\bf F}_U
 - \delta_i^{-2}\,\widehat{\bf F}_T, \\[0.5ex]
\frac{3}{2}\frac{d_i \hat{p}_i}{d\hat{t}} + \frac{5}{2}\,
\hat{p}_i\,\widehat{\nabla}\cdot\widehat{\bf V}_i + \delta_i^{2}\,
\zeta_i^{-1}\,\widehat{\bpi}_i : \widehat{\nabla}\cdot\widehat{\bf V}_i
+\zeta_i^{-1} \,\widehat{\nabla}\cdot\widehat{\bf q}_{i}
&= \delta_i^{-2}\,\zeta_i\,\mu\,\widehat{w}_i. \label{e3.116c}
\end{align}
The only large or small (compared to unity) quantities in these equations are the
parameters $\delta_i$, $\zeta_i$, and $\mu$. 
Here, $d_i/d\hat{t}\equiv\partial /\partial\hat{t} +
 \widehat{\bf V}_i\cdot\widehat{\nabla}$.

@@

\subsection{Electromagnetic Torques}\label{storque}
The flux-surface integrated poloidal and toroidal electromagnetic torque densities acting on the plasma
can be written
\begin{align}\label{e152}
T_\theta(r) &= \left\langle r\,{\bf j}\times {\bf B}\cdot {\bf e}_\theta\right\rangle\\[0.5ex]
T_\varphi(r) &= \left\langle R_0\,{\bf j}\times {\bf B}\cdot {\bf e}_z\right\rangle,
\end{align}
respectively, 
where
\begin{equation}
\left\langle\cdots\right\rangle \equiv \oint\oint r\,R_0\,(\cdots)\,d\theta\,d\varphi
\end{equation}
is a flux-surface integration  operator. 
 However, according to Equation~(\ref{e80}),
\begin{equation}\label{e150}
{\bf j}\times {\bf B} \simeq \nabla p.
\end{equation}
Given that the scalar pressure is a single-valued function of $\theta$ and $\varphi$, it immediately follows
that $T_\theta=T_\varphi = 0$ throughout the plasma (Fitzpatrick 1993). The only exception to this rule occurs in the immediate vicinity of the
resonant surface, where Equation~(\ref{e150}) breaks down. It follows that we can write
\begin{align}\label{e156a}
T_\theta(r) &= T_{\theta\,s}\,\delta(r-r_s),\\[0.5ex]
T_\varphi(r) &= T_{\varphi\,s}\,\delta(r-r_s),
\end{align}
where
\begin{align}\label{e153}
T_{\theta\,s} &=  \frac{1}{4}\int_{r_{s-}}^{r_{s+}}\oint\oint R_0\,r^{\,2}\,(\delta j_z\,\delta B_r^{\,\ast} + \delta j_z^{\,\ast}\,\delta B_r)\,dr\,d\theta\,d\varphi,\\[0.5ex]
T_{\varphi\,s} &=  -\frac{1}{4}\int_{r_{s-}}^{r_{s+}}\oint\oint R_0^{\,2}\,r\,(\delta j_\theta\,\delta B_r^{\,\ast} + \delta j_\theta^{\,\ast}\,\delta B_r)\,dr\,d\theta\,d\varphi
\end{align}
are the net poloidal and toroidal torques, respectively, acting at the resonant surface. 
Note that the zeroth-order (in perturbed quantities) torques are zero because $B_r=0$. Furthermore, the linear (in perturbed quantities) torques 
average to zero over the flux-surface. Hence, the largest non-zero torques  are {\em quadratic}\/ in perturbed quantities. 
According to Equation~(\ref{e82}), 
\begin{equation}
\nabla \cdot \delta {\bf j} = {\rm i}\left(\frac{m}{r}\,\delta j_\theta - \frac{n}{R_0}\,\delta j_z\right) = 0,
\end{equation}
where $\delta{\bf j}$ is the density of the current sheet flowing at the resonant surface, and we have made use
of the fact that $\delta j_r\simeq 0$ in a radially thin current sheet. 
 It follows from the previous three equations that
\begin{equation}\label{e156}
T_{\varphi\,s} = - \frac{n}{m}\,T_{\theta\,s}.
\end{equation}
Finally, Equations~(\ref{e93}), (\ref{e94}), (\ref{e153}), and (\ref{e156}) imply that
\begin{align}\label{e157}
T_{\theta\,s} &= -\frac{2\pi^2\,R_0\,m}{\mu_0}\,{\rm Im}({\mit\Delta\Psi}_s\,{\mit\Psi}_s^{\,\ast}),\\[0.5ex]
T_{\varphi\,s} &= \frac{2\pi^2\,R_0\,n}{\mu_0}\,{\rm Im}({\mit\Delta\Psi}_s\,{\mit\Psi}_s^{\,\ast})\label{e158}
\end{align}
(Fitzpatrick 1993).

@@

Equations~(\ref{e182}) and (\ref{e185}) are subject to the boundary conditions
\begin{align}\label{e186}
\frac{\partial{\mit\Delta\Omega}_\theta(0,t)}{\partial r} &= \frac{\partial{\mit\Delta\Omega}_\varphi(0,t)}{\partial r} =0,\\[0.5ex]
{\mit\Delta\Omega}_\theta(a,t) &= {\mit\Delta\Omega}_\varphi(a,t)=0.\label{e187}
\end{align}
The boundary conditions (\ref{e186}) merely ensure that the ion angular velocities remain finite at the magnetic axis. 
On the other  hand, the boundary conditions (\ref{e187}) are a consequence of the action of {\em charge exchange}\/ with electrically neutral particles emitted isotropically from the wall  in the edge regions of the plasma
(Brau, et al., 1983; Fitzpatrick 1993; Monier-Garbet, et al., 1997). Charge exchange with neutrals gives rise to 
dominant damping torques acting at the edge of the plasma that relax the edge ion angular velocities  toward  particular values. Moreover, the
electromagnetic torques that develop at the resonant surface are not large enough, compared with the charge-exchange torques,
to significantly modify the edge ion angular velocities (Fitzpatrick 1993).

@@

Equations~(\ref{e220})--(\ref{e223}) are known collectively as the {\em drift-MHD equations}, and contain both shear-Alfv\'{e}n
and compressible-Alfv\'{e}n wave dynamics. The system of equations is completed by Maxwell's equations:
\begin{align}
\nabla\cdot{\bf B} &= 0,\\[0.5ex]
\nabla\times {\bf E} &= - \frac{\partial {\bf B}}{\partial t},\\[0.5ex]
\mu_0\,{\bf j} &= \nabla \times {\bf B}.\label{e226}
\end{align}
